The Implications of Local Fluctuations in the Galactic Midplane for Dynamical Analysis in the \textit{Gaia} Era

### 2019-08-22 paper/ms.tex
Fix previous commit
Update KVJ's NSF grant
Remove bold from ref comments

\textbf{The true Galactic potential is not strictly axisymmetric, beyond even
the typically-quoted gas clouds, spiral arms, and bar(s). The presence of disk
oscillations has been known since its observation in H~{\sc i} by
\citet{1957AJ.....62...93K}. The presence of a North-South asymmetry in
stellar number density and velocity in the solar neighborhood has been found
in SDSS data \citep{2012ApJ...750L..41W}, and in velocity in the LAMOST survey
\citep{2013ApJ...777L...5C, 2015ApJ...801..105X}, and RAVE data
\citep{2013MNRAS.436..101W}. Beyond the solar neighborhood, low-latitude
overdensities have been observed, e.g., the Monoceros Ring
\citep{2002ApJ...569..245N, 2003MNRAS.340L..21I, 2003ApJ...594L.119C,
2014ApJ...791....9S, 2016ApJ...825..140M}, an overdensity in the direction of
the Triangulum and Andromeda galaxies \citep{2007ApJ...668L.123M,
2014ApJ...793...62S, 2015ApJ...801..105X, 2015MNRAS.452..676P} and A13
\citep{2010ApJ...722..750S, 2017ApJ...844...74L}, and in the direction of the
Galactic Center \citep{2014Natur.509..342F}. There is evidence that these
stellar populations originated from the disk \citep{2015MNRAS.452..676P,
2018ApJ...854...47S, 2018Natur.555..334B}. See, e.g.,
\citet{2018MNRAS.481..286L} for a discussion of these observations.}

\textbf{With the vast improvement in the quality of phase-space measurements
due to \textit{Gaia} the assumption of axisymmetry is increasingly inadequate
\citep[e.g.,][]{2018Natur.561..360A, 2019MNRAS.485.3134L}. The presence of the
phase-space spiral itself is evidence of disk oscillations
\citep{2018Natur.561..360A}, but the fact that its strength changes across the
disk is further evidence \citep{2019MNRAS.485.3134L, 2019MNRAS.486.1167B}.
Even if the axisymmetric assumption were close enough for many purposes, the
parameters used in axisymmetric models of the Galactic potential may be biased
by the non-axisymmetries in the disk.}

@@

\textbf{In reality one only ever measures the position and velocity of a star
at a single instance in time. However, there are at least two possibilities
for going beyond a single temporal snapshot of the phase-space distribution.
If more than one star on the same or similar orbit could be identified (e.g.,
by identifying the remnants of a disrupted open cluster), then it would be
possible to constrain the disk potential in analogy to how stellar streams are
used to constrain the Galactic  potential \citep[e.g.][]{1999ApJ...512L.109J,
2001ApJ...551..294I, 2004ApJ...610L..97H, 2015ApJ...801...98S,
2015ApJ...803...80K, 2019MNRAS.483.1427L}. This would be challenging since, in
principle, one would want to use the stellar actions or other conserved
quantities, but as we discuss in this work they are subject to substantial
errors.  Another possibility would be to use chemical tagging to identify a
disrupted open cluster \citep{2002ARA&A..40..487F}, but this may be difficult
due to the large number of stars that are chemically indistinguishable though
presumably born in different clusters \citep{2018ApJ...853..198N}. A second
way to improve on a single temporal snapshot is to measure directly the
accelerations of stars \citep{2018arXiv181207581S}. While this would allow a
direct measurement of the aliasing effect in Figure~\ref{fig:cartoon}, the
observational challenges will likely limit this type of measurement to a small
number of local stars for the near future.}

@@

\textbf{Because the energies in the radial and vertical coordinates are
related to the radial and vertical actions by a constant
(Equation~\eqref{eq:JR_Jz_epi}), the fractional error in these two energies is
the same as the actions.\footnote{This ignores any potential error in the
epicyclic and vertical frequencies.} The energy in the azimuthal coordinate
can be readily computed ($E_{\phi} = \frac{1}{2}v_{\phi}^2$). Using standard
propagation of errors, we see that $\Delta E_{\phi}/E_{\phi}$ and $\Delta
J_{\phi}/J_{\phi}$ (Equation~\eqref{eq:induced_Jphi}) differ by a $\Delta
R/R_g$ term. Since $R_g\approx 8\,\kpc$, we expect this term to be negligible
and we therefore conclude that:
\beq\label{eq:error_total_energy}
\frac{\Delta E}{E} \approx \frac{\Delta J_R}{J_R} + 2\frac{\Delta J_{\phi}}{J_{\phi}}
                     + \frac{\Delta J_z}{J_z}\text{.}
\eeq}

@@

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report $\Delta
J_i/J_i$ for the three different fiducial orbits in Table~\ref{tab:orbits}.
The upper panels of Figure~\ref{fig:many_orbit_wrong_ref} shows the spread
induced in each action for an offset in the $z$-component. In the lower panels
we consider offsets in the $x$ component (i.e. the solar radius). \textbf{The
left, center, and right columns show the fractional spread in the values
computed for $J_R$, $J_{\phi}$, and $J_z$, respectively.}

@@

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For \thin{}, a systematic offset of $15\,\pc$ in the $z$-coordinate
results in a $25\%$ deviation in $J_z$, while a $120\,\pc$ offset results in
the same deviation for \thick{}. We find that \halo{} is relatively
\textbf{insensitive} to errors in the midplane, with only $\sim15\%$ error in
$J_z$ out to an offset of $500\,\pc$.

@@

\subsection*{Supplementary Material}
All code used in this work is available at
\url{https://github.com/gusbeane/actions_systematic} and archived at Zenodo
[doi:10.5281/zenodo.3364082].

\acknowledgments
We thank the anonymous referee for providing helpful comments. We would like
to thank Megan Bedell, Robert A. Benjamin, Maria Bergemann, Joss
Bland-Hawthorn, Tobias Buck, Elena D'Onghia, Benoit Famaey, and Adrian
Price-Whelan for helpful discussions. A.B. would like to thank Todd Phillips
for helpful discussions. This work uses data hosted by the Flatiron
Institute's FIRE data hub. The Flatiron Institute is supported by the Simons
Foundation. This project was developed in part at the 2019 Santa Barbara Gaia
Sprint, hosted by the Kavli Institute for Theoretical Physics (KITP) at the
University of California, Santa Barbara. This research was supported in part
at KITP by the Heising-Simons Foundation and the National Science Foundation
(NSF) under Grant No. PHY-1748958. This work used the Extreme Science and
Engineering Discovery Environment (XSEDE), which is supported by NSF Grant No.
OCI-1053575. A.B. was supported in part by the Roy \& Diana Vagelos Program in
the Molecular Life Sciences and the Roy \& Diana Vagelos Challenge Award.
K.V.J.'s contributions were supported in part by the NSF under Grant No.
AST-1614743. M.-M.M.L. was partly supported by the NSF under Grant No.
AST-1815461.

>>>>

The true Galactic potential is not strictly axisymmetric, beyond even
the typically-quoted gas clouds, spiral arms, and bar(s). The presence of disk
oscillations has been known since its observation in H~{\sc i} by
\citet{1957AJ.....62...93K}. The presence of a North-South asymmetry in
stellar number density and velocity in the solar neighborhood has been found
in SDSS data \citep{2012ApJ...750L..41W}, and in velocity in the LAMOST survey
\citep{2013ApJ...777L...5C, 2015ApJ...801..105X}, and RAVE data
\citep{2013MNRAS.436..101W}. Beyond the solar neighborhood, low-latitude
overdensities have been observed, e.g., the Monoceros Ring
\citep{2002ApJ...569..245N, 2003MNRAS.340L..21I, 2003ApJ...594L.119C,
2014ApJ...791....9S, 2016ApJ...825..140M}, an overdensity in the direction of
the Triangulum and Andromeda galaxies \citep{2007ApJ...668L.123M,
2014ApJ...793...62S, 2015ApJ...801..105X, 2015MNRAS.452..676P} and A13
\citep{2010ApJ...722..750S, 2017ApJ...844...74L}, and in the direction of the
Galactic Center \citep{2014Natur.509..342F}. There is evidence that these
stellar populations originated from the disk \citep{2015MNRAS.452..676P,
2018ApJ...854...47S, 2018Natur.555..334B}. See, e.g.,
\citet{2018MNRAS.481..286L} for a discussion of these observations.

With the vast improvement in the quality of phase-space measurements
due to \textit{Gaia} the assumption of axisymmetry is increasingly inadequate
\citep[e.g.,][]{2018Natur.561..360A, 2019MNRAS.485.3134L}. The presence of the
phase-space spiral itself is evidence of disk oscillations
\citep{2018Natur.561..360A}, but the fact that its strength changes across the
disk is further evidence \citep{2019MNRAS.485.3134L, 2019MNRAS.486.1167B}.
Even if the axisymmetric assumption were close enough for many purposes, the
parameters used in axisymmetric models of the Galactic potential may be biased
by the non-axisymmetries in the disk.

@@

In reality one only ever measures the position and velocity of a star
at a single instance in time. However, there are at least two possibilities
for going beyond a single temporal snapshot of the phase-space distribution.
If more than one star on the same or similar orbit could be identified (e.g.,
by identifying the remnants of a disrupted open cluster), then it would be
possible to constrain the disk potential in analogy to how stellar streams are
used to constrain the Galactic  potential \citep[e.g.][]{1999ApJ...512L.109J,
2001ApJ...551..294I, 2004ApJ...610L..97H, 2015ApJ...801...98S,
2015ApJ...803...80K, 2019MNRAS.483.1427L}. This would be challenging since, in
principle, one would want to use the stellar actions or other conserved
quantities, but as we discuss in this work they are subject to substantial
errors.  Another possibility would be to use chemical tagging to identify a
disrupted open cluster \citep{2002ARA&A..40..487F}, but this may be difficult
due to the large number of stars that are chemically indistinguishable though
presumably born in different clusters \citep{2018ApJ...853..198N}. A second
way to improve on a single temporal snapshot is to measure directly the
accelerations of stars \citep{2018arXiv181207581S}. While this would allow a
direct measurement of the aliasing effect in Figure~\ref{fig:cartoon}, the
observational challenges will likely limit this type of measurement to a small
number of local stars for the near future.

@@

Because the energies in the radial and vertical coordinates are
related to the radial and vertical actions by a constant
(Equation~\eqref{eq:JR_Jz_epi}), the fractional error in these two energies is
the same as the actions.\footnote{This ignores any potential error in the
epicyclic and vertical frequencies.} The energy in the azimuthal coordinate
can be readily computed ($E_{\phi} = \frac{1}{2}v_{\phi}^2$). Using standard
propagation of errors, we see that $\Delta E_{\phi}/E_{\phi}$ and $\Delta
J_{\phi}/J_{\phi}$ (Equation~\eqref{eq:induced_Jphi}) differ by a $\Delta
R/R_g$ term. Since $R_g\approx 8\,\kpc$, we expect this term to be negligible
and we therefore conclude that:
\beq\label{eq:error_total_energy}
\frac{\Delta E}{E} \approx \frac{\Delta J_R}{J_R} + 2\frac{\Delta J_{\phi}}{J_{\phi}}
                     + \frac{\Delta J_z}{J_z}\text{.}
\eeq

@@

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report $\Delta
J_i/J_i$ for the three different fiducial orbits in Table~\ref{tab:orbits}.
The upper panels of Figure~\ref{fig:many_orbit_wrong_ref} shows the spread
induced in each action for an offset in the $z$-component. In the lower panels
we consider offsets in the $x$ component (i.e. the solar radius). The
left, center, and right columns show the fractional spread in the values
computed for $J_R$, $J_{\phi}$, and $J_z$, respectively.

@@

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For \thin{}, a systematic offset of $15\,\pc$ in the $z$-coordinate
results in a $25\%$ deviation in $J_z$, while a $120\,\pc$ offset results in
the same deviation for \thick{}. We find that \halo{} is relatively
insensitive to errors in the midplane, with only $\sim15\%$ error in
$J_z$ out to an offset of $500\,\pc$.

@@

\subsection*{Supplementary Material}
All code used in this work is available at
\url{https://github.com/gusbeane/actions_systematic} and archived at Zenodo
[doi:\href{https://doi.org/10.5281/zenodo.3364082}{10.5281/zenodo.3364082}].

\acknowledgments
We thank the anonymous referee for providing helpful comments. We would like
to thank Megan Bedell, Robert A. Benjamin, Maria Bergemann, Joss
Bland-Hawthorn, Tobias Buck, Elena D'Onghia, Benoit Famaey, and Adrian
Price-Whelan for helpful discussions. A.B. would like to thank Todd Phillips
for helpful discussions. This work uses data hosted by the Flatiron
Institute's FIRE data hub. The Flatiron Institute is supported by the Simons
Foundation. This project was developed in part at the 2019 Santa Barbara Gaia
Sprint, hosted by the Kavli Institute for Theoretical Physics (KITP) at the
University of California, Santa Barbara. This research was supported in part
at KITP by the Heising-Simons Foundation and the National Science Foundation
(NSF) under Grant No. PHY-1748958. This work used the Extreme Science and
Engineering Discovery Environment (XSEDE), which is supported by NSF Grant No.
OCI-1053575. A.B. was supported in part by the Roy \& Diana Vagelos Program in
the Molecular Life Sciences and the Roy \& Diana Vagelos Challenge Award.
K.V.J.'s contributions were supported in part by the NSF under Grant No.
AST-1715582. M.-M.M.L. was partly supported by the NSF under Grant No.
AST-1815461.
